Columnar VCF for Biobank scale datasets using Zarr

### 2024-04-18 paper.tex

The data shown in Figure~\ref{fig-whole-matrix-compute} for Zarr and Savvy
is based on custom programs written using their respective APIs
to implement the \texttt{af-dist} operation. The Zarr program uses
the Zarr-Python package to iterate over the decoded chunks of the 
genotype matrix and classifies genotypes within a chunk using a 14 line Python
function, accelerated using the Numba JIT compliler~\cite{lam2015numba}.
The allele frequencies and genotype counts are then analysed to produce 
the final counts within the allele frequency bins with 9 lines of 
Python using NumPy~\cite{harris2020array} functions. Remarkably, this 
short and simple Python program is substantially faster than the 
equivalent compiled C using \texttt{htslib} APIs on BCF (6 hours
vs 20 hours for 1 million samples). The fastest method is the 
C++ program written using the Savvy API. This would largely seem
to be due to Savvy's excellent genotype decoding performance
% zarr 1.2 GiB
% savvy 6.6 GiB
% zarr_nshf 3.9 GiB
(up to 6.6GiB/s vs 1.2GiB/s for Zarr on this dataset;
Figure~\ref{fig-whole-matrix-decode}).
% sequence_length  num_samples  num_sites       tool       size
% 40         48129895      1000000    7254858       zarr  22.071687
% 35         48129895      1000000    7254858        bcf  51.749294
% 36         48129895      1000000    7254858    genozip  10.691384
% 37         48129895      1000000    7254858        sav  21.249436
% 38         48129895      1000000    7254858        tsk   1.802636
% 39         48129895      1000000    7254858        vcf  81.375831
% 41         48129895      1000000    7254858  zarr_nshf  29.897540
Turning off the BitShuffle filter for the Zarr dataset,
however, leads to a substantial increase in decoding speed
(3.9GiB/s) at the cost of a roughly 25\% increase in storage
space (29.9GiB up from 22.1GiB for 1 million samples; data not
shown). Given the relatively small contribution of genotypes to the
overall storage of real datasets (see the Genomics England example)
and the frequency that they are likely to be accessed, this
would seem like a good tradeoff in most cases.
This ability to easily tune compression performance
and decoding speed on a field-by-field basis is a major strong
point of Zarr. The \texttt{vcf2zarr} utility also provides
functionality to aid with such storage schema tuning.

>>>>

The data shown in Figure~\ref{fig-whole-matrix-compute} for Zarr and Savvy
is based on custom programs written using their respective APIs
to implement the \texttt{af-dist} operation. The Zarr program uses
the Zarr-Python package to iterate over the decoded chunks of the 
genotype matrix and classifies genotypes within a chunk using a 14 line Python
function, accelerated using the Numba JIT compliler~\cite{lam2015numba}.
The allele frequencies and genotype counts are then analysed to produce 
the final counts within the allele frequency bins with 9 lines of 
Python using NumPy~\cite{harris2020array} functions. Remarkably, this 
short and simple Python program is substantially faster than the 
equivalent compiled C using \texttt{htslib} APIs on BCF (6.9 hours
vs 20.6 hours for 1 million samples). 
% num_samples  num_sites      tool  user_time  sys_time     wall_time  total_time
%     1000000    7254858     savvy    8371.07      4.16   8377.718136  2.326453
%     1000000    7254858  bcftools   73939.32     46.09  74023.083678  20.551503
%     1000000    7254858      zarr   24709.64     29.32  24750.704171  6.871933
The fastest method is the 
C++ program written using the Savvy API. This would largely seem
to be due to Savvy's excellent genotype decoding performance
% zarr 1.2 GiB
% savvy 6.6 GiB
% zarr_nshf 3.9 GiB
(up to 6.6GiB/s vs 1.2GiB/s for Zarr on this dataset;
Figure~\ref{fig-whole-matrix-decode}).
% sequence_length  num_samples  num_sites       tool       size
% 40         48129895      1000000    7254858       zarr  22.071687
% 35         48129895      1000000    7254858        bcf  51.749294
% 36         48129895      1000000    7254858    genozip  10.691384
% 37         48129895      1000000    7254858        sav  21.249436
% 38         48129895      1000000    7254858        tsk   1.802636
% 39         48129895      1000000    7254858        vcf  81.375831
% 41         48129895      1000000    7254858  zarr_nshf  29.897540
Turning off the BitShuffle filter for the Zarr dataset,
however, leads to a substantial increase in decoding speed
(3.9GiB/s) at the cost of a roughly 25\% increase in storage
space (29.9GiB up from 22.1GiB for 1 million samples; data not
shown). Given the relatively small contribution of genotypes to the
overall storage of real datasets (see the Genomics England example)
and the frequency that they are likely to be accessed, this
would seem like a good tradeoff in most cases.
This ability to easily tune compression performance
and decoding speed on a field-by-field basis is a major strong
point of Zarr. The \texttt{vcf2zarr} utility also provides
functionality to aid with such storage schema tuning.
